Debugging by \textit{lastChange}

### 2010-10-19 papers/pldi2011/sigplanconf-template.tex

As a proof of concept, we developed \textit{Querypoint}, a prototype which enhances Firebug JavaScript debugger with \textit{lastChange} feature. Moreover, \textit{Querypoint} provides mechanisms for automated bug reproduction, and a novel user interface to investigated execution points and results.

@@

Our contribution in this paper is the algorightm \textit{lastchange}, which locates the last place a value is changed, gathers other values from that execution point, and allows \textit{lastChange} operations from that point. The algorithm builds on existing breakpoint debugger technology. We demonstrate the feasibiltily of the algorithm with \textit{Querypoint}, an implementation extending the Firebug JavaScript debugger. \textit{Querypoint} also provides mechanism for automated bug reproduction, and a centralized user interface which summarizes investigated execution points and collected results.

@@

\subfigure[A view of Firebug JavaScript debugger. The selected script panel, presents all loaded JavaScript code. Developer has access to any loaded source file and can set breakpoints on lines. In this figure, the execution is paused at line 22 by a regular breakpoint. The panel at the right site shows the program state at the paused point. ]{\label{fig:example1}\includegraphics[width=1.0\textwidth, height=.19\textheight]{1-bp22.jpg}}

@@

The \textit{lastChange} algorithm is based on program re-execution. Developer sets a breakpoint and execution is paused once the breakpoint hits. The breakpoint hit becomes the \textit{Reproduction point} when developer asks for the \textit{lastChange} of a value. Debugger re-executes the program, collects limited data and shows the result. \textit{Reproduction point} is the execution point in which first \textit{lastChange} is calld. Like the introductory example, it can be defined by a [conditional] breakpoint. \textit{lastChange} can be called on two different types of values: An object property or a variable value. In JavaScript, every available variable in a frame comes from one of the scopes in the current scope chain. There are five different scope types: global, local, with, catch, closure. We explain each case separately.

\subsection{\textit{lastChange} on object propery}
To simplify the algorithm explanation and avoid technical details, we assume two basic operations and postpone implementation details to the next section. The first opertion is \texttt{objectId()} which gets an object and returns an integer as its identifier. This identifier is unique for every object during one execution. The second operation is \texttt{setPropertyChangeHook()} which gets function as the hook and a string as the property name like \texttt{foo}. Whenever property \texttt{foo} in any object changes the hook will be called. The hook function receives a reference to the owner of \texttt{foo}.

Once developer asks for the last change of \texttt{bar.foo} at the reproduction point, debugger calls \texttt{setPropertyChangeHook()} with \texttt{foo} as the property name and re-executes the program. Whenever \texttt{foo} changes and the hook function is called, debugger first calls \texttt{objectID()} on the \texttt{foo} owner object. Then it stores owner id and stack frame locations. The reason we keep object id instead of a reference to the object itself is to let garbage collector to destroy dead objects. Whenever the execution reaches the reproduction point it looks at the history of \texttt{foo} changes and finds the last \texttt{foo} change with the same object id as \texttt{bar} id at the reproduction point. Figure ~\ref{fig:foo-changes1} shows the list of property \texttt{foo} change events in a hypothetical execution. \texttt{bar} id at the reproduction point is 1010, so the last change of \texttt{bar.foo} is the forth column.

@@

\subsection{\textit{lastChange} on variable} 
In JavaScript, every frame has a scope chain and every available variable in the frame comes from one of the scopes in the frame's scope chain. Once developer asks for the last change of variable \texttt{foo} at the reproduction point, debugger first determines the variable's scope as follows: it iterates over the scopes in the scope chain and the first scope which has a variable with the same name is the variable's scope. There are five different scope types: local, global, \texttt{with}, closure, \texttt{catch}. We explain each case separately.

\subsubsection{global and \texttt{with} scopes}
Global scope is the the most outer scope in the scope chain and it also refers as global object. This scope is a regular JavaScript object and therefore every global variable is a property of global object. Similar to global scope, \texttt{with} scopes are also regular objects. A \texttt{with} scope is created by a \texttt{with()} block with an object as the parameter. Every property of passed object is available inside the block as a variable. Debugger treats the case where variable's scope is global or \texttt{with}, like \textit{lastChange} on an object property.

\subsubsection{local, closure and \texttt{catch} scopes}
local scope refers to the most nested scope in the scope chain. Clousure scope refers to the scope which is created once a function is defined. Catch scope is the scope created by try-catch blocks and contains the exception variable. These objects are not necessarily regular JavaScript objects. Therefore to track changes to a variable in these scopes we need a new strategy.

%%% we need a short explanation about scopes
%%% important: think about it: we also need to set a special variable (id) to let us to identify the related functional scope
\begin{figure}[htp]
\begin{verbatim}
10  function parent(arg1){
11    var var1;
12    var1 = 0;
13    function myfun(arg2){
14      var var2;
15      var2 = {...}
16      var1 = var2;
17      function firstChild(arg3){
18        var1 = 1;
19        var2 = 1;
20      }  
21      function secondChild(arg4){
22       	var var1;
23        var1 = 2			      
24        var2 = 2;	
25      }
26    }  
27  }    
\end{verbatim}
\caption{Sample JavaScript code demonstrating closure variables.}
\label{fig:js-closure}
\end{figure}

A variable in these scopes can only be changed by the code which is included in the its definition block. Therefore it is only enough to take care only about the lines that a variable is assigned a new value. The same fact is also true about \texttt{catch} and local scopes. Debugger puts hooks in on the line the variable is defined and all lines the variable is assigned a new value (including inner-functions). For example in Figure ~\ref{fig:js-closure}, if the execution is paused at line 19, four hooks on lines 11, 12, 16, and 18, but if the execution is paused at line 24 and the last change of \texttt{var1} is requested by the developer, two hooks on lines 22 and 23 will be set. The reason is that \texttt{var1} in \texttt{firstChild} is a closure variable but in \texttt{secondChild} is a local variable.

@@

Recursive function: 
 f(){
   var x;
   x++;
   f();
   x; 
   ...
 }

The trace of recursive calls:

  f was called.
    x changes; // A
      f was called;
        x changes;
        f was called 
          x changes; //C
    //C

\end{verbatim}
\caption{Recursive calls.}
\label{fig:recursive}
\end{figure}

@@

Everytime one of the hook hit a new event is inserted to the list of events. To separate irrelvant event from each other, we use a special variable \texttt{\_scopeId}. If the scope has this variable, the value of this variable is used for the scope id otherwise a new id will be generated and \texttt{\_scopeId} variable is created by \texttt{eval()} function in the scope. For example evaluating \texttt{var \_scopeId = 50} will create a variable with name \texttt{\_scopeId} and value \texttt{50} in the local scope. Consider that even in the case of closure scopes, \texttt{\_scopeId} is set when the variable first defined and the variable's scope is still local. Figure ~\ref{fig:recursive} demonstrates one case that scope id is necessary for identifying the right answer. In thiss case the same variable \texttt{x} is changed several times but in different scopes due to recursive calls. Without the scope id calling \textit{lastChange(x)} at point C will return point B while the right answer is point A.

@@

\begin{center}
\textit{
 point A : the reproduction point \\
 point B : lastChange(A, bar.x) \\
 point C : lastChange(B, baz.y) 
 }
 \end{center}
According the algorithm explained in the previous section none of the points B and C can be identified before reaching point A. Therefore, all \texttt{x} changes for piont B and all \texttt{y} changes for point C are stored. Once the execution reaches point A, and the exact \texttt{bar} id reveals, point B can be identified among the items in the history of \texttt{x} changes. The problem arises in recognizing point C. To recognize point C, we need \texttt{baz} id at point B, but B is a past program state, and \texttt{baz} id has not been collected at this point.

@@

To address this issue, debugger does a dependency analysis between all \textit{lastChange} queries and computes the additional data needs to be collected at each point. For example in the mentioned case, \texttt{baz} id- if it is available- should be collected at all points \texttt{x} changes. Having this additional data, locating point C, using the \texttt{baz} id at point B is straightforward (Figure ~\ref{fig:lastchange-lastchange}).

>>>>

As a proof of concept, we developed \textit{Querypointer}, a prototype which enhances Firebug JavaScript debugger with \textit{lastChange} feature. Moreover, \textit{Querypointer} provides mechanisms for automated bug reproduction, and a novel user interface to investigated execution points and results.

@@

Our contribution in this paper is the algorightm \textit{lastchange}, which locates the last place a value is changed, gathers other values from that execution point, and allows \textit{lastChange} operations from that point. The algorithm builds on existing breakpoint debugger technology. We demonstrate the feasibiltily of the algorithm with \textit{Querypointer}, an implementation extending the Firebug JavaScript debugger. \textit{Querypointer} also provides mechanism for automated bug reproduction, and a centralized user interface which summarizes investigated execution points and collected results.

@@

\subfigure[A view of Firebug JavaScript debugger. The selected script panel, presents all loaded JavaScript code. Developer has access to all loaded source file and can set breakpoints on lines. In this figure, the execution is paused at line 22 by a regular breakpoint. The panel at the right site shows the program state at the paused point. ]{\label{fig:example1}\includegraphics[width=1.0\textwidth, height=.19\textheight]{1-bp22.jpg}}

@@

The story starts when developer examines the program state at a breakpoint hit and asks for the \textit{lastChange} of a value. The \textit{lastChange} algorithm is based on program re-execution. Therefore it is assumed that the execution can be reproduced automatically (e.g., by a given test case) or with the developer's assistance. The breakpoint hit becomes the \textit{reproduction point}. Debugger sets hooks on all instructions that might be the result of \textit{lastChange} query, re-executes the program, at every hook hit, \textit{change event}, collects limited data and once the execution reaches the reproduction point, it analyzes the collected data and shows the result. The result is a partial snapshot of the program state at the execution point that last change occured.

As we shown in the previous section, a \textit{lastChange} query can be performed on the result of another \textit{lastChange} query. If we name the reproduction point \textit{R}, we can write the first \textit{lastChange} in the introductory example in this form: \textit{lastChange(R, myObject.myProperty)}. It means that this query is defined at \textit{R}. If we name the result of this query \textit{A}, we can write the second \textit{lastChange} in this form: \textit{lastChange(A, myCondition.value)}. In this way, a sequence of \textit{lastChange} queries with any length can be defined.

\textit{lastChange} can be called on two different types of values: An object property or a variable value. In JavaScript, every available variable in a frame comes from one of the scopes in the frame's scope chain. There are five different scope types: global, local, with, catch, closure. We explain each case separately.

\subsection{\textit{lastChange} on object propery}
To simplify the algorithm explanation and avoid technical details, we assume two basic operations and later we explain the details of this two operations. The first opertion is \texttt{objectId()} which gets a JavaScript object and returns an integer as its identifier. This identifier is unique for every object during one execution. The second operation is \texttt{setPropertyChangeHook()} which gets function as the callback and a string as the property name like \texttt{foo}. Whenever property \texttt{foo} in any object changes the callback will be called. The callback function receives a reference to the owner of \texttt{foo}.

@@

Once developer asks for the last change of \texttt{bar.foo} at the reproduction point, debugger calls \texttt{setPropertyChangeHook()} with \texttt{foo} as the property name and re-executes the program. Whenever \texttt{foo} changes and the callback function is called, debugger first calls \texttt{objectId()} on the \texttt{foo} owner object. Then it stores owner id and stack frame locations. The reason we keep object id instead of a reference to the object itself is to let garbage collector to destroy dead objects. Whenever the execution reaches the reproduction point it looks at the history of \texttt{foo} changes and finds the last \texttt{foo} change with the same object id as \texttt{bar} id at the reproduction point. Figure ~\ref{fig:foo-changes1} shows the list of property \texttt{foo} change events in a hypothetical execution. \texttt{bar} id at the reproduction point is 1010, so the last change of \texttt{bar.foo} is the forth column.

\subsubsection{\texttt{objectId()}}
The object id is kept as a property of the object, \texttt{\_objectid}, which is not enumerable. Therefore, it will not appear in \texttt{for(property in object)} loops and has no effect on the executing program. Once \texttt{objectId()} is called on an object and the object has not \texttt{\_objectId} property, this property is defined by calling \texttt{defineProperty} JavaScript function. As a parameter it can be specified that the defined property is not enumerable. Then it is enough to assign a new generated id to this property.

\subsubsection{\texttt{setPropertyChangeHook()}}
To implement this operation, we use two features that are available in Firefox JavaScript engine. First, every regular JavaScript object has a \texttt{watch()} function which receives two parameters, a property name and another function as a callback. Whenever the property with the given name changes the calleback function is called. The hook is set by this function is enabled even if the property is removed and added again [ref]. The issue with this function is that it can only be applied to available objects. However, at the beginning of re-execution there is no object available. To solve this issue we use the second feature. Enabling a flag gives us the place the object created including the JavaScript file url and line number (e.g., myFile.js, line 24). Though this is a limited data, it helps us to get a reference to the object once it is created. Then it is enough to set objet id and call \texttt{watch()} with the property name. In section 4.1, we explain how we get reference to the created objects.

\subsection{\textit{lastChange} on variable} 
In JavaScript, every frame has a scope chain and every available variable in the frame comes from one of the scopes in the frame's scope chain. Once developer asks for the last change of variable \texttt{foo} at the reproduction point, debugger first determines the variable's scope as follows: it iterates over the scopes in the scope chain and the first scope which has a variable with the same name is the variable's scope. There are five different scope types: local, global, \texttt{with}, closure, \texttt{catch}. We explain these cases in two groups.

\subsubsection{global and \texttt{with} scopes}
Global scope is the the most outer scope in the scope chain and it is also refered as global object which is usually \texttt{window} object. This scope is a regular JavaScript object and therefore every global variable is a property of global object. Similar to global scope, \texttt{with} scopes are also regular objects. A \texttt{with} scope is created by a \texttt{with()} block with an object as the parameter. Every property of passed object is available inside the block as a variable. Debugger treats the case where variable's scope is global or \texttt{with}, like \textit{lastChange} on an object property.

\subsubsection{local, closure and \texttt{catch} scopes}
Local scope refers to the most nested scope in the scope chain. Closure scope refers to the scope which is created once a function is defined. Catch scope is the scope created in the catch block of try-catch statements and contains the exception variable. These scopes are not necessarily regular JavaScript objects. Therefore, to track changes to a variable in these scopes we use a different strategy.

\begin{figure}[htp]
\begin{verbatim}
10  function parent(){
11    var var1;
12    var1 = ...;
13    function myfun(){
14      var1 = ...;
15      function firstChild(){
16       	var var1;
17        var1 = ...;
18      }  
19      function secondChild(){
20        var1 = ...;			      
21      }
22    }  
23  }    
\end{verbatim}
\caption{Sample JavaScript code demonstrating local and clousure variables.}
\label{fig:js-closure}
\end{figure}

A variable in these types of scope can only be changed in the function the variable is defined or the nested fuctions. For example in Figure ~\ref{fig:js-closure}, variable \texttt{var1} in line 11 can only be changed in lines 10 to 23. Therefore, after locating the function in which the variable is defined, it is enough to parse the code inside the function block and set a hook on all lines where the variable is assigned a new value.% TODO lines or instruction???
We also set a hook at the first line of the function which is corresponding to the line variable is defined. In Figure ~\ref{fig:js-closure}, if the execution is paused at line 17 and the last change of \texttt{var1} is queried, two hooks on lines 16 and 17 will be set, but if the execution is paused at line 20 , four hooks on lines 11, 12, 14, 20 will be set. These two cases are different due the fact that \texttt{var1} in \texttt{firstChild} is a local variable but in \texttt{secondChild} is a closure variable.

Everytime one of the hooks hits a new change event is inserted to the list of change events. During execution the same variable might be changed several times but in different scopes. Figure ~\ref{fig:recursive} demonstrates one case that a scope identifier is necessary for locating the right answer. In this case the same variable \texttt{x} is changed several times but in different scopes due to the recursive calls. Without the scope id calling \textit{lastChange(x)} at point C will return point B while the right answer is point A.

@@

  function f(){
    var x;
    x++;
    f();
  }

Sample Trace:

   f()
A  |  x changes; 
   |  f()
   |  |  x changes;
   |  |  f()
B  |  |  |  x changes; 
C  |  ?

\end{verbatim}
\caption{A recursive call trace demonstrates the need for scope id. Querying \textit{lastChange} on \texttt{x} at point C should return point A not B.}
\label{fig:recursive}
\end{figure}

@@

The scope id is kept as a variable in the scope, \texttt{\_scopeId}. As we said, a hook is set on the first instruction of the function defines the variable. Whenever this hook hits,  \texttt{\_scopeId} variable is set by \texttt{eval()} function. For example evaluating \texttt{var \_scopeId = 50} will create a variable with name \texttt{\_scopeId} and value \texttt{50} in the local scope.

@@

\begin{center}
\textit{
 point A : the reproduction point \\
 point B : lastChange(A, bar.x) \\
 point C : lastChange(B, baz.y) 
 }
 \end{center}
According to the algorithm explained in the previous section none of the points B and C can be identified before reaching point A. Therefore, all \texttt{x} changes for piont B and all \texttt{y} changes for point C are stored. Once the execution reaches point A, and the exact \texttt{bar} id reveals, point B can be identified among the change events in the history of \texttt{x} changes. The problem arises in recognizing point C. To recognize point C, we need \texttt{baz} id at point B, but B is a past program state, and \texttt{baz} id has not been collected at this point.

To address this issue, debugger does a dependency analysis between all \textit{lastChange} queries and computes the additional data required to be collected at each point. For example in the mentioned case, \texttt{baz} id- if it is available- should be collected at all points \texttt{x} changes. Having this additional data, locating point C, using \texttt{baz} id at point B is straightforward (Figure ~\ref{fig:lastchange-lastchange}).

\begin{algorithm}
\caption{\textit{lastChange} queries dependency analysis.}
\label{dependency-analysis}
\begin{algorithmic}

\FOR{$q$ in $queries$} 
 \FOR {$p$ was defined at the $q$ result}
   \IF {$p$ is a lastChange on object property} 
     \STATE property id should be stored at $q$ change events. 
   \ELSIF {$p$ is a lastChange on variable} 
     \STATE variable scope id should be stored at $q$ change events.
	 \ENDIF 
 \ENDFOR 
\ENDFOR

\end{algorithmic}
\end{algorithm}

@@
